
title: "M05-1: Principles - Expanded In-depth Data Wrangling with Tidyverse in R"
author: "Paola Garcia"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 4
    number-sections: true
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
    df-print: paged
editor: visual
execute: 
  freeze: auto
  warning: false  
  error: true
---

# Introduction

In this workshop, we learn how to wrangle data using the **Tidyverse** packages in R.  
The most commonly used packages for data wrangling include **dplyr** and **tidyr**.  
These packages help transform, organize, and clean data so it is easier to analyze.

# Load Required Packages

```{r}
library(tidyverse)
```

# Example Dataset

For demonstration purposes, we will use the built-in **mtcars** dataset.

```{r}
head(mtcars)
```

# Tidying Data with tidyr

## pivot_longer()

The `pivot_longer()` function converts data from **wide format to long format**.

```{r}
library(tidyr)

long_data <- mtcars %>%
  pivot_longer(
    cols = mpg:hp,
    names_to = "variable",
    values_to = "value"
  )

head(long_data)
```

## pivot_wider()

The `pivot_wider()` function converts data from **long format back to wide format**.

```{r}
wide_data <- long_data %>%
  pivot_wider(
    names_from = variable,
    values_from = value
  )

head(wide_data)
```

## separate()

The `separate()` function splits a column into multiple columns.

```{r}
df <- data.frame(name = c("John-Doe", "Jane-Smith"))

separate(df, name, into = c("first", "last"), sep = "-")
```

## unite()

The `unite()` function combines multiple columns into one.

```{r}
df2 <- data.frame(
  first = c("John","Jane"),
  last = c("Doe","Smith")
)

unite(df2, "full_name", first, last, sep = " ")
```

# Transforming Data with dplyr

## select()

The `select()` function chooses specific columns from a dataset.

```{r}
mtcars %>%
  select(mpg, cyl, hp)
```

## filter()

The `filter()` function keeps rows that meet a condition.

```{r}
mtcars %>%
  filter(cyl == 6)
```

## arrange()

The `arrange()` function sorts rows in a dataset.

```{r}
mtcars %>%
  arrange(desc(mpg))
```

## mutate()

The `mutate()` function creates a new column based on existing data.

```{r}
mtcars %>%
  mutate(mpg_km = mpg * 0.425)
```

## group_by()

The `group_by()` function groups the dataset for analysis.

```{r}
mtcars %>%
  group_by(cyl)
```

## count()

The `count()` function counts observations by group.

```{r}
mtcars %>%
  count(cyl)
```

# Example of Grouped Analysis

We can combine multiple functions together.

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = mean(mpg),
    avg_hp = mean(hp)
  )
```

# Conclusion

The **Tidyverse** provides powerful tools for data wrangling and transformation.  
Functions from **tidyr** help tidy and restructure datasets, while functions from **dplyr** allow us to filter, select, and transform data efficiently.  
These techniques are essential for preparing data before performing analysis or visualization.